A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-01-18 analysis/paper/paper.Rmd
large update to paper generation. Includes supplementary figures at the end. Fix references. Text and figure update

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
author:
  - Wouter Saelens* ¹ ²
  - Robrecht Cannoodt* ¹ ² ³
  - Helena Todorov¹ ² ⁴ ⁵
  - Yvan Saeys¹ ²
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: nature-biotechnology.csl
abstract: |
  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Dozens of trajectory inference methods have been developed, early methods mostly focussing on linear trajectories, while recently methods also model more complex branching trajectories. However, the performance of these methods have not yet been comprehensively evaluated. Here, we compare a total of 22 trajectory inference methods, on both realistic synthetic datasets and a large set is real datasets. We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended, and is available at www.github.com/dynverse.
keywords: |
  trajectory inference; pseudotemporal ordering; single-cell transcriptomics
highlights: |
  These are the highlights. 
---
\* Equal contribution  
¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.  
⁵ Université Claude Bernard Lyon 1, Lyon, France.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = ".scratch/"
)

@@

  if (anchor) ref <- pritt("{ref}<a name='{ref_type}_{ref_id}'></a>")
  ref
}
create_names <- list(
  sfig = function(i) pritt("Supplementary Figure {i}"),
  fig = function(i) pritt("Figure {i}"),
  snote = function(i) pritt("Supplementary Note {i}")
)
```
<!----
Tips for working in this document:
* Go to tools -> perferences and remove Use smart quotes
* Add enters after every paragraph (this is markdown!)

@@

```{r, fig.width = 15, fig.height = 10}
n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_id="method_characteristics"))
trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_id="method_characteristics"))

cowplot::plot_grid(plotlist=list(n_methods_over_time, trajectory_components_over_time), nrow=2, rel_heights = c(0.7, 0.3), labels="auto")
```

`r ref("sfig", "methods_over_time", anchor=TRUE)` New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.

@@

We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(method_df$wrapper == "Done")` for evaluation, primarily based on their free availability and the presence of a programming interface. We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and underlying algorithm(s) (`r ref("fig", "method_characteristics")`).

@@

While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further. We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("snote", "qc_scoring")`). The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or for which purpose(s) they are important (user friendliness, developer friendliness or good science).

After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation. Most methods reach a decent score, fulfilling most of the quality criteria (`r ref("fig", "method_characteristics", "b")`) recent methods tend to have higher code quality sFIG. Several aspects are lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation are problematic areas in particular (`r ref("sfig", "method_qc_check_plot")` and `r ref("fig", "method_characteristics", "c")`).

@@

![](`r figure_file("figure_methods.svg", experiment_id="method_characteristics")`)

`r ref("fig", "method_characteristics", anchor=TRUE)` First characterization of each method.

```{r, fig.width = 10, fig.height = 13}
readRDS(figure_file("method_qc_check_plot.rds"))
```
`r ref("sfig", "method_qc_check_plot", anchor=TRUE)`

```{r, fig.width = 15, fig.height = 7.5}
readRDS(figure_file("method_components_plot.rds"))
```
`r ref("sfig", "method_components", anchor=TRUE)` TI methods frequently share steps

### Evaluation overview

![evaluation_overview](../../analysis/figures/evaluation_overview.png)
`r ref("fig", "evaluation_overview", anchor=TRUE)` Overview of the evaluation

@@

>>>>

---
title: "A comparison of single-cell trajectory inference methods: towards more accurate and robust tools"
author:
  - Wouter Saelens* ¹ ²
  - Robrecht Cannoodt* ¹ ² ³
  - Helena Todorov¹ ² ⁴ ⁵
  - Yvan Saeys¹ ²
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
    fig_caption: yes
    reference_docx: templates/template.docx
bibliography: references.bib
csl: nature-biotechnology.csl
abstract: |
  Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Dozens of trajectory inference methods have been developed, early methods mostly focussing on linear trajectories, while recently methods also model more complex branching trajectories. However, the performance of these methods have not yet been comprehensively evaluated. Here, we compare a total of 22 trajectory inference methods, on both realistic synthetic datasets and a large set is real datasets. We employ a novel parameter optimisation procedure, and compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended, and is available at www.github.com/dynverse.
keywords: |
  trajectory inference; pseudotemporal ordering; single-cell transcriptomics
highlights: |
  These are the highlights. 
---
\* Equal contribution  
¹ Data mining and modelling for biomedicine, VIB Center for Inflammation Research, Ghent, Belgium.  
² Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium.  
³ Center for Medical Genetics, Ghent University Hospital, Ghent, Belgium.  
⁴ CIRI, Centre International de Recherche en Infectiologie, Inserm, Lyon, France.  
⁵ Université Claude Bernard Lyon 1, Lyon, France.  
<!----
Tips for working in this document:
* Go to tools -> perferences and remove Use smart quotes
* Add enters after every paragraph (this is markdown!)

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = paste0(rprojroot::find_rstudio_root_file(), "/analysis/paper/.scratch/")
)

@@

  if (anchor)  ref <- pritt("{ref}{anchor(ref_type, ref_id)}")

  ref
}
anchor <- function(ref_type, ref_id) {
  pritt("<a name='{ref_type}_{ref_id}'></a>")
}
create_names <- list(
  sfig = function(i) pritt("Supplementary Figure {i}"),
  fig = function(i) pritt("Figure {i}"),
  snote = function(i) pritt("Supplementary Note {i}")
)

sfigs <- tibble(ref_id = character(), fig = list(), caption = character(), width=numeric(), height=numeric())
add_sfig <- function(fig, ref_id, caption, width=15, height=10) {
  sfigs <<- sfigs %>% add_row(
    fig = list(fig), 
    ref_id = ref_id, 
    caption=caption, 
    width=width, 
    height=height
  )
}
```

@@

```{r}
add_sfig({
n_methods_over_time <- readRDS(figure_file("n_methods_over_time.rds", experiment_id="method_characteristics"))
trajectory_components_over_time <- readRDS(figure_file("trajectory_components_over_time.rds", experiment_id="method_characteristics"))

cowplot::plot_grid(plotlist=list(n_methods_over_time, trajectory_components_over_time), nrow=2, rel_heights = c(0.7, 0.3), labels="auto")
}, 
"methods_over_time", 
"New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory.
", 
15, 
10
)
```

@@

We gathered a list of `r sum(method_df$is_ti == "Yup", na.rm=TRUE)` TI methods from literature, and selected a subset of `r sum(method_df$wrapper == "Done")` for evaluation, primarily based on their free availability and the presence of a programming interface. We characterized all methods in four different ways: possible trajectory structures, implementation quality, prior information and the underlying algorithm (`r ref("fig", "method_characteristics")`).

@@

While not directly related to the accuracy of the inferred trajectory, the quality of the implementation is an important first evaluation metric, because good unit testing makes sure the implementation is correct, good documentation makes it easier for potential users to apply the method on their data, and overall good code quality makes it possible for other developers to adapt the method and extend it further. We therefore looked at the implementation of each method, and assessed their quality using a transparent scoring scheme (`r ref("snote", "qc_scoring")`). The individual quality checks can be grouped in two ways: what aspect of the method they investigate (availability, code quality, code assurance, documentation, behaviour and the paper) or for which purpose(s) they are important (user friendliness, developer friendliness or good science). After publishing this preprint, we will contact the authors of each method, allowing them to improve their method before the final publishing of the evaluation.

Most methods fulfilling most of the basic criteria, such as free availability and basic code quality criteria (`r ref("fig", "method_characteristics", "b")`). Recent methods tend to have higher code quality sFIG. However, several aspects are consistently lacking for the majority of the methods (`r ref("fig", "method_characteristics", "e")`) and we believe that these should receive extra attention from developers. Although these outstanding issues cover all five categories, code assurance and documentation in particular are problematic areas (`r ref("sfig", "method_qc_check")` and `r ref("fig", "method_characteristics", "c")`), notwithstanding several studies pinpointing these as good practices [@wilson_best_2014].

@@

`r anchor("fig", "method_characteristics")`
![](`r figure_file("figure_methods.svg", experiment_id="method_characteristics")`)
`r ref("fig", "method_characteristics", anchor=TRUE)` First characterization of each method.

```{r}
add_sfig({
readRDS(figure_file("method_qc_check_plot.rds", experiment_id="method_characteristics"))
}, 
"method_qc_check", 
"Caption", 
10, 
13
)
```

```{r}
add_sfig({
readRDS(figure_file("method_components_plot.rds", experiment_id="method_characteristics"))
}, 
"method_components", 
"TI methods frequently share steps", 
15, 
7.5
)
```
### Evaluation overview
`r anchor("fig", "evaluation_overview")`
![evaluation_overview](../../analysis/figures/evaluation_overview.png)
`r ref("fig", "evaluation_overview")` Overview of the evaluation

@@

# Supplementary Figures

```{r, echo=FALSE, results='asis'}
subchunkify <- function(ref_id, caption, width=5, height=7) {
  subchunk <- glue::glue("",
  "",
  "`r anchor('sfig', '{ref_id}')`",
  "```{{r {ref_id}, fig.height={height}, fig.width={width}, echo=FALSE}}",
  "sfigs %>% filter(ref_id == !!ref_id) %>% pull(fig) %>% .[[1]]",
  "```",
  "`r ref('sfig', '{ref_id}')` {caption}",
  "", .sep = "\n")

  cat(knitr::knit(text=subchunk, quiet=TRUE))
}

pwalk(sfigs, function(ref_id, fig, caption, width, height) {
  subchunkify(ref_id, caption, width, height)
})
```
